

Question 4 Part A : Bases
Question 4 Part B : Linear Independence
Question 4 Row Space and Column Space
Question 5 : Gram Scmidt Process : Projection Operator
Question 5 part B : Least Squares Approximation
 
 
System of Linear Equations
A system of linear equations can be written in matrix form: Ax=b
 
A is typically an  nn matrix of coefficients
 
x is a n1 column of unknown variables , which the system seeks to find solutions for.
 
b is typically an  n1 matrix of "solution" coefficients
 
The system can expressed in terms of its transposr
 
xtAt=bt(Ax)t=(b)t
 
The system is only solvable if Det(A) 0
Question 4 Part A : Bases
In linear algebra, a basis is a set of vectors in a given vector space with certain properties:

 
One can get any vector in the vector space by multiplying each of the basis vectors by different numbers, and then adding them up. 
If any vector is removed from the basis, the property above is no longer satisfied. 
 

 
The dimension of a given vector space is the number of elements of the basis.
 
 
All quadratic expressions (i.e.px2+ qx+r ) can be written as linear combination of those terms.
 
i.e. For any given set of values{p,q,r} there exists some coefficients {a,b,c} such that
 
px2+ qx+r = a(1-2x) + b(2 +x2) +c(4x+x2)
 
 
1) We can write the above equation in matrix form
 
            
 
        px2+ qx+r = 
 
 
 = 
 
  
2) simplifying
arraycp & q & rarray = 
 
 
3) This is a system of linear equations. ( in form xtAt=bt, as opposed to Ax=b )  
 
 
This system is only solvable if  is invertible
 
i.e.det(C) 0
 
4) Lets check it.
 
Using the first column
  
det(C) = 1(01) -1(41) - 2(-21) -1(40) + 0
 
 
det(C) = -4 +4 +0 = 0
 
 
Determinant is zero!
 
The system of equations is not solvable.
 
{(1-2x),(2 +x2),(4x+x2)} is not a basis in R2
 
 
 
Question 4 Part B : Linear Independence
Determine whether the following set of vectors in R4 is linearly independent.
 
     
 
If one column can be expressed as a linear combination of the other columns, then they are not linearly independent
 
We will use an inspection based approach and then by computing the determinant of the aggregation.
 
 
Inspection 
 
Consider the last column v4. We are going to see if it can be expressed as a linear combination of the other columns.
 
If it can, we can compute {a,b,c }, where v4=av1+bv2+cv3
 
Necessarily a must equal 3.
 
Let us compute the difference v4-av1=v4-3v1
 
      = 
 
This difference is equivalent 3v2
 
 
Necessarily the last column vector is a linear combination of the other vectors v4=3v1+3v2 
 
Formal approach
 
Choose one of the columns ( we pick v4 )
 
Try to find {a,b,c }, where v4=av1+bv2+cv3
If we cant solve this - then they are linearly independent.
 
 
      = 
 
Cell 1 : a=3
 
Cell 4: -3=-a+c	c=0
 
Do we get consistent solutions from cells 2 and 3
Cell 2 :0=2a-2b	a=b	b=3
 
Cell 3 : -3=a-2b  b=3  
 
 
Necessarily the last column vector is a linear combination of the other vectors 
 
v4=3v1+3v2+0v3 
 
 
 
 
  
 
 
 
